% =====================================================================
% background.tex  --  "Polarization Defusing" project
% Drop this into your Overleaf project and \input{} it, or paste the body.
% Requires in your preamble:
%   \usepackage{booktabs}      % \toprule \midrule \bottomrule
%   \usepackage{amssymb}       % \checkmark
%   \usepackage{tabularx}      % wide comparison table
% Bibliography: \bibliography{refs} (or add refs.bib to your project).
% Citation commands assume natbib/ACM \cite{}; adjust if you use \citep/\citet.
% =====================================================================

\newcommand{\yes}{\checkmark}
\newcommand{\no}{--}
\newcommand{\prt}{$\sim$}   % partial / indirect

\section{Background and Related Work}
\label{sec:background}

Our project asks how to \emph{defuse} polarization in large interpersonal
online groups---recruiting already-polarized communities and intervening in
their everyday chat medium (forum threads, group messaging, Discord-style
channels) with a population of generative-AI agents that exert \emph{social
influence} rather than top-down moderation. This goal sits at the intersection
of three lines of work that our lab and the broader HCI community have
developed: (i)~how opinion and behavior polarize in online communities,
(ii)~how large language models (LLMs) and multi-agent systems can carry social
influence, and (iii)~how conversational and playful interventions shift
attitudes. Table~\ref{tab:rw} positions representative prior systems---five of
which are from our own lab---against the design commitments of the present
project, and the three subsections below unpack each line in turn.

% ---------------------------------------------------------------------
\begin{table*}[t]
  \centering
  \footnotesize
  \setlength{\tabcolsep}{4pt}
  \renewcommand{\arraystretch}{1.2}
  \caption{Refined map of related work against the design commitments of
    \emph{Polarization Defusing}. \yes~= central, \prt~= partial/indirect,
    \no~= absent. ``Target'' marks whether the work moves \emph{opinion} (O) or
    \emph{behavior} (B). Bottom row is the gap our project fills.}
  \label{tab:rw}
  \begin{tabularx}{\textwidth}{@{}>{\raggedright\arraybackslash}p{2.7cm} >{\raggedright\arraybackslash}X c c c >{\raggedright\arraybackslash}p{2.6cm} c c@{}}
    \toprule
    \textbf{Prior work} & \textbf{Everyday medium} & \textbf{Reduce polariz.} & \textbf{LLM/ GenAI} & \textbf{Multi-agent} & \textbf{Influence lever} & \textbf{Target} & \textbf{Play} \\
    \midrule
    Ronaldo's a poser~\cite{zeng2025ronaldo}      & Online forum (sports debate) & \prt & \yes & \no  & Argument construction        & O   & \no  \\
    Can AI Prompt Humans~\cite{zhang2025ecoecho}  & Serious game (NPC dialogue)  & \no  & \yes & \yes & Prompt act + show consequence& B   & \yes \\
    Breaking the News~\cite{tang2025breaking}     & Social-media-sim game        & \prt & \yes & \yes & Role-play + inoculation      & O   & \yes \\
    Eternagram~\cite{zhou2024eternagram}          & Chat / text adventure        & \prt & \yes & \no  & Narrative immersion          & O   & \yes \\
    Hear You in Silence~\cite{jiang2026hear}      & 1:1 conversational agent     & \no  & \yes & \no  & Active listening / pacing    & \prt& \no  \\
    Govers et al.~\cite{govers2024aimediators}    & Online debate                & \yes & \yes & \no  & AI mediation                 & O   & \no  \\
    Bad News~\cite{roozenbeek2019badnews}         & Social-media-sim game        & \prt & \no  & \no  & Inoculation                  & O   & \yes \\
    \midrule
    \textbf{Ours (Polariz. Defusing)} & \textbf{Group chat / forum / Discord} & \textbf{\yes} & \textbf{\yes} & \textbf{\yes} & \textbf{Peer social influence} & \textbf{O\,/\,B} & \textbf{\prt} \\
    \bottomrule
  \end{tabularx}
\end{table*}

% =====================================================================
\subsection{Polarization, Opinion Dynamics, and Debate in Online Communities}
\label{sec:rw-polarization}

Online communities form around heterogeneous goals---entertainment,
information exchange, social support, and prestige---and their early
trajectories are shaped by the motivations of their founders and the bonds
among members~\cite{kairam2024communities}. When such communities organize
around rival identities, however, contact between groups does not necessarily
breed tolerance. Characterizing NBA discussion forums, \citet{zhang2019sportshate}
show that high intergroup contact correlates with markedly more negative and
profane language, and studies of online football communities trace how
emotionally charged expression escalates into aggression, trolling, and
downvoting spirals~\cite{wang2023emotional}. Crucially, the dynamics of
polarization are subtler than the ``echo chamber'' intuition suggests:
\citet{efstratiou2022nonpolar} find that hostility is not simply a product of
cross-group exposure, and in some settings in-group members are \emph{more}
polarized than those who cross group lines, while \citet{marchal2022polarization}
documents that interactions between ideologically opposed users are
significantly more negative than like-minded ones---an affective, identity-driven
form of polarization. These findings motivate our decision to recruit
\emph{already-polarized} groups and to treat polarization as both an opinion
and an affective phenomenon, rather than assuming mere exposure will defuse it.

A complementary thread studies how people actually argue. Persuasion in
good-faith online debate has measurable interaction dynamics and strategy
signatures~\cite{tan2016winning}, and classical argumentation theory supplies
structure---claims, grounds, warrants, and rebuttals~\cite{toulmin2003uses}---as
well as catalogs of the fallacies (ad hominem, straw man, hasty generalization)
that corrode reasoning~\cite{vaneemeren2014handbook}. Our own
\emph{Ronaldo's a poser!}~\cite{zeng2025ronaldo} brings these threads into the
generative-AI era, observing how soccer fans recruit ChatGPT to construct
arguments in a live forum and how their rhetoric shifts when new participants
arrive. That work characterizes the \emph{problem}---GenAI amplifying
competitive, polarized debate---and directly sets up the question our project
inverts: can a population of agents instead be designed to \emph{defuse} it?

% =====================================================================
\subsection{LLMs and Multi-Agent Systems for Social Influence and Behavior Change}
\label{sec:rw-multiagent}

If polarization is sustained through social interaction, then so might its
remedy be. Generative agents can convincingly simulate human behavior,
maintaining memory, reflection, and plausible social action at
scale~\cite{park2023generative}, which makes a \emph{population} of LLM agents a
viable substrate for staging social influence inside a community rather than
addressing users one at a time. Our \emph{Can AI Prompt Humans?}~\cite{zhang2025ecoecho}
demonstrates the inversion of the usual human-prompts-AI relationship:
multimodal agents prompt players' actions and surface their consequences,
significantly shifting intended behavior---evidence that agent-initiated
influence can move people toward, in that case, sustainable conduct. Most
directly relevant to defusing, \citet{govers2024aimediators} design AI-driven
\emph{mediation} strategies for audience depolarization in online debates, and
\citet{tanprasert2024filterbubble} show that debate chatbots can foster critical
thinking and broaden perspectives, with social identity and conversational
style modulating their effect. These results suggest that the \emph{stance} and
\emph{social positioning} of an AI interlocutor, not merely its arguments,
govern whether it depolarizes.

Why social positioning matters is explained by persuasion theory. The
Persuasion Knowledge Model holds that people actively recognize and resist
perceived persuasion attempts~\cite{friestad1994persuasion}, and
heuristic--systematic accounts show that source cues such as perceived
expertise are weighed alongside argument quality~\cite{reimer2004hsm}.
Top-down, overtly corrective messaging therefore risks triggering reactance in
exactly the polarized users we target. This is the central rationale for our
multi-agent design: rather than a single authoritative debunker, a community of
peer-like agents can exert lateral social influence---modeling moderate
positions, normalizing disagreement, and reducing the salience of the
persuasion attempt itself. Our \emph{Breaking the News}~\cite{tang2025breaking}
prototypes part of this machinery, using LLM agents to simulate a reactive
public whose ``opinion'' players try to sway, demonstrating that LLM-driven
social audiences can be both believable and consequential.

% =====================================================================
\subsection{Conversational Agents and Playful Interventions for Attitude Change}
\label{sec:rw-intervention}

The final question is how an agent should \emph{talk} so that polarized users
stay engaged rather than entrench. A substantial HCI literature on
conversational agents shows that perceived listening is foundational:
interview chatbots equipped with active-listening skills improve user
experience~\cite{xiao2020ifihear}, backchanneling elicits self-disclosure and
strengthens the sense of being heard~\cite{cho2022alexa}, and chatbots designed
for reciprocal disclosure encourage users to open up in
turn~\cite{lee2020ihear}. Social Penetration Theory frames why this
matters---relationships, and the trust that makes attitude change possible,
deepen through graduated mutual disclosure~\cite{altman1973social}---while the
Computers Are Social Actors paradigm establishes that people respond to
sufficiently social agents with the same scripts they apply to
humans~\cite{reeves1996media,nass2000machines}, so that agent pacing and
listening cues carry real social weight. Our \emph{Hear You in Silence}~\cite{jiang2026hear}
extends this to context-aware pacing and strategic silence, showing that
\emph{how} and \emph{when} an agent responds shapes human-likeness, trust, and
disclosure---design levers we intend to give our defusing agents so that they
earn standing to influence before they ever challenge a view.

Finally, playful and inoculation-based framings offer a low-reactance route to
attitude change. Inoculation games such as \emph{Bad News} confer psychological
resistance to misinformation by letting players experience weakened forms of
manipulation~\cite{roozenbeek2019badnews}, an effect analyzed through
inoculation and transportation theory across a family of such
games~\cite{grace2023prebunking}. Our lab has repeatedly used this register:
\emph{Eternagram}~\cite{zhou2024eternagram} probes and nudges climate attitudes
through a ChatGPT-driven text adventure, and \emph{Breaking the News} and
\emph{Can AI Prompt Humans} (above) embed attitude and behavior shifts in game
mechanics. This grounds the optional \emph{play} dimension of our project: a
forum or Discord channel can host light gamified or role-play elements that make
encountering an opposing view feel exploratory rather than confrontational.

% =====================================================================
\paragraph{Summary and gap.}
Across these three lines, prior work has separately established that online
groups polarize through affective, identity-driven interaction
(\S\ref{sec:rw-polarization}); that LLM and multi-agent systems can credibly
simulate people and steer opinion or behavior, with mediation and peer framing
mattering more than raw argument (\S\ref{sec:rw-multiagent}); and that
listening, disclosure, and play make AI interlocutors trusted and
non-threatening enough to shift attitudes (\S\ref{sec:rw-intervention}). As
Table~\ref{tab:rw} shows, no existing system combines all of our commitments:
intervening in an \emph{everyday} group medium, targeting \emph{already-polarized}
populations, and using a \emph{multi-agent} community of GenAI peers to defuse
\emph{either} opinion or behavioral polarization through social influence. Our
project occupies precisely this gap.
