% =====================================================================
% introduction.tex  --  "Polarization Defusing" (Lakers vs Celtics)
% Draft Introduction. Hands off into \input{background}.
% Citation keys live in refs.bib.
% =====================================================================

\section{Introduction}
\label{sec:intro}

Affective polarization---warmth toward one's in-group paired with animus toward
the out-group, largely independent of actual positional disagreement---has become
a defining feature of online life~\cite{iyengar2015fear,iyengar2019origins}. It is
fundamentally a matter of group identity rather than issue content: categorizing
people into groups is by itself enough to generate in-group favoritism and
out-group hostility~\cite{tajfel1979sit}. On the platforms where people now spend
ordinary social time---forums, group chats, and social feeds---this identity-driven
antagonism is enacted as toxic, self-reinforcing discourse. Our own study of a live
fan forum shows rivals recruiting generative AI to sharpen their attacks and
escalating as opposing newcomers arrive~\cite{zeng2025ronaldo}, and synthetic
social networks populated by LLM agents reliably reproduce these polarized dynamics
under controlled conditions~\cite{donkers2025synthetic}. The recurring pattern is a
pathology of large interpersonal groups: the more rival identities talk past one
another, the further apart they move.

Existing responses are poorly matched to this dynamic. Content moderation,
fact-checking, and corrective messaging treat polarization as a problem of bad
information delivered to individuals, yet overt persuasion provokes psychological
reactance---a rebound toward the very position it targets~\cite{brehm1966reactance}.
Recent HCI evidence sharpens the warning for AI specifically: when an AI simply
agrees with or validates a user's stance, confidence in the polarized position
\emph{increases} and depolarization becomes less likely~\cite{whenaiagrees2026}.
These interventions also operate one user at a time, against a phenomenon that is
fundamentally \emph{social}---sustained by group identity, peer signaling, and the
felt presence of an audience. If generative AI can amplify polarized
debate~\cite{zeng2025ronaldo}, the inverse question follows: can a \emph{population}
of AI agents instead be arranged to defuse it?

We propose \emph{polarization defusing}: a simulated social-media platform that
recruits members of an already-polarized rivalry and intervenes through a population
of generative-AI agents exerting peer-level \emph{social influence} rather than
top-down correction. Our intervention is \emph{structural} rather than rhetorical.
Drawing on the Common Ingroup Identity Model~\cite{gaertner2000reducing} and the
finding that superordinate cooperative goals reduce intergroup
conflict~\cite{sherif1961robbers,pettigrew2006contact}, we arrange the platform so
that rival fans \emph{discover} a shared, more inclusive identity (``we are all NBA
fans'') through cooperative cross-group interaction---a recategorization shown to
shift even rival sports fans toward helping one another~\cite{levine2005identity}.
The agents do not argue positions; instead, multiple peer-like agents \emph{model}
de-escalated behavior and inclusive turn-taking for participants to imitate,
establishing a calmer descriptive norm. Our prior systems supply the building
blocks: multimodal agents that prompt human action~\cite{zhang2025ecoecho}, LLM
agents that simulate a reactive public~\cite{tang2025breaking}, and conversational
agents whose pacing and active listening earn social standing~\cite{jiang2026hear}.

We study the Los Angeles Lakers versus Boston Celtics rivalry: a historic,
city-and-identity-based conflict that is widely recognized across both Western and
Asian audiences, making it a tractable yet ecologically familiar model of affective
polarization. Because Social Identity Theory and the Common Ingroup Identity Model
are domain-general, an intervention that works here speaks to affective
polarization more broadly. This design lets us study defusing along two axes prior
work has treated separately: whether the intervention shifts \emph{opinion}
(expressed stance, affect toward the out-group) or \emph{behavior} (how members
post, escalate, and engage). Accordingly, we ask:
\begin{itemize}
  \item \textbf{RQ1 (process / behavior).} How do polarized fan groups interact
  and engage within a simulated social-media platform, and what behavioral changes
  emerge when a structured recategorization mechanic is introduced?
  \item \textbf{RQ2 (effect).} How does a Common-Ingroup-Identity--based,
  multi-agent LLM intervention embedded in the platform affect affective
  polarization between the opposing fan groups?
  \item \textbf{RQ3 (mechanism).} Is any reduction in polarization mediated by
  participants' perceived common identity, and does it persist after the
  intervention is withdrawn?
\end{itemize}

This work makes the following anticipated contributions: (1)~the framing and design
of \emph{polarization defusing} as a structural, multi-agent social-influence
intervention sited in a simulated social platform; (2)~a system in which peer-like
GenAI agents drive cooperative recategorization and model de-escalated norms; and
(3)~empirical findings, on the Lakers--Celtics rivalry, on whether and how such
agents defuse affective and behavioral polarization. The remainder of the paper
develops three lines of background---the sociological, psychological, and technical
foundations of this approach---which we review next.
